%\section{The von Neumann architecture}

An alternative programming paradigm, the \key{imperative}, is historically
closer to hardware. It emerged as the programming paradigm for computer
architectures stemming from John von Neumann's original proposal for the design
of the digital computer\cite{von-neumann}.  A key aspect of the von Neumann
architecture is that it keeps the computer rather feebleminded, leaving it to
the programmer to be the clever one. Inevitably, no good introduction to the
imperative programming paradigm can commence, without at least a gentle
introduction to the von Neumann architecture.

\begin{definition}

A \key{computer}\footnotemark is an entity with \key{state} and
\key{processors}. A processor performs \key{actions} that change the state of
the computer.

\footnotetext{Although at the beginning of the 20th century, a ``computer'' was
still a profession, today most computers are electronic machines.  We'll use
the words ``computer'' and ``machine'' interchangably.}

\end{definition}

While it may be modern (about time) to speak of computers as machines with
multiple processors, we will restrain ourselves to single processor machines.
Programming for multiple processors raises a range of problematics irrelevant
to the core matter of these lecture notes. At times it will provide for an
interesting discussion to consider how certain aspects map over into
multiprocessor machines, but in such cases they will be mentioned explicitly.

\begin{definition}

``An action is a happening, taking place in a finite period of time and
establishing a well-defined, intended \key{net effect}.''
\cite{dijkstra-introduction}

\end{definition}

This definition highlights two important points. Firstly, an action takes place
in a finite period of time, say $T_1$. This allows us to speak of the points in
time $T_n$ and $T_{n+1}$, as the times of action inception and termination,
i.e. $T_{n+1}=T_n+T_1$. Secondly, an action establishes a well-defined,
intended net effect. This highlights that we are interested in actions that do
something, indeed something expected.

To this end, it makes sense to describe the ``net effect'' of an action as the
difference between the state of the computer at time $T_n$ and $T_{n+1}$.
However, this notion breaks down as we turn to multiprocessor machines, where
the clear benefit of performing multiple actions at once has been utilized.
Apply this notion with care.

% ``action – a function or a function object that mutates the value of an
% object'' -- Alexander Stepanov, Adobe notes.

As an example of an action, consider doing the laundry. Given a pile of dirty
laundry, the net effect of this action is the same pile of laundry, but clean.
Clearly, the functional paradigm fits abnormally to this type of problem.
Knitting a pile of clean laundry, same but different, is not nearly as
practical as cleaning the dirty pile. Indeed, why should we dispose of a pile
of perfectly good clothing, just because it has gotten a little dirty? The
process of cleaning, rather than knitting, is explicit to the programmer in the
imperative paradigm.

While modifying global state is a marvel of imperative programming, it is
important that we achieve an intended net effect; nothing more, nothing less.
So while we wish to change the state of the world in some particular way, we
also wish to avoid unintended \key{side-effects}.  This is to avoid a general
``butterfly effect'', where doing the laundry causes a thunderstorm on the
other side of the globe.

% A bit too early for this, but worth noting in general: ``The great discovery
% of Turing and von Neumann that set us on a new path was the discovery of
% memory.  We are not just dealing with numbers: we are storing them in
% different locations.''\cite{stepanov-adobe}}

% A programming language facilitates a programming paradigm if it is both easy,
% and reasonably safe to use the paradigm.

One technique that the imperative paradigm shares with the functional is
\key{data abstraction}. The idea is to group the elements of the world into
well-defined \key{data structures} and modify data structures rather than
arbitrary global state.

\begin{definition}

A data structure is a particular ordering of data elements. A data structure
has well-defined methods of modification that maintain this ordering.

\end{definition}

The ``data elements'' that a data structure maintains are data structures
themselves. So the methods of modification of a particular data structure have
to take into account the methods of modification of its constituent data
elements.

This technique allows for arbitrary data abstraction, but of course, at some
point, some actual modification of the global state has to commence. A computer
therefore has  at least one \key{basic} data structure, whos methods of
modification are implemented in hardware. We'll refer to these basic methods of
modification as the computer's \key{instructions}. The state of the computer
can then be a finite \key{array} of instances of this basic data structure. We
call this array the computer's \key{memory}.

\begin{definition}

An array is a collection of data elements. Every element has a unique
\key{index} that identifies its position within the array. An array has
\key{random access}.

\end{definition}

An array is a data structure. For all intents and purposes it does not matter
\emph{how} the elements are actually ordered. We would simply like to be able
to \key{access} the elements in arbitrary order and modify their values. For
this purpose we require that the data structure possesses ``random access'',
which is to say that regardless of which element we attempt to access, it will
take roughly the same amount of time.

One frequently used basic data structure is a \key{word}: a fixed size group of
bits, typically 4 bytes in size. For example:

\begin{equation}
0000 0000\ 0000 0000\ 0000 0000\ 0000 0000 \label{word-0}
\end{equation}

A possible basic method of modification is ``flip all bits'', so that we can go
back and forth between the above and the below:

\begin{equation}
1111 1111\ 1111 1111\ 1111 1111\ 1111 1111 \label{word-1}
\end{equation}

Such a data structure may not seem particularly useful, but it \emph{can}
represent a boolean value. An $n$-size array of such values could for example
be used to automatically monitor a parking lot with $n$ parking spots, where
each spot is either ``free'' or ``occupied''. Imagine that we have an auxiliary
sensor system that notifies us whenever a car enters or leaves a particular
spot. We can then flip all bits of a corresponding word whenever either even
occurs. However, for such a system to operate consistently this instruction is
not enough.

Initially, the parking lot is completely empty. Regardless of whether we chose
to use \ref{word-0} or \ref{word-1} to represent an ``empty'' spot, we cannot
consistently model this reality. We have no guarantee that a particular word is
initially either one value or the other. For such purposes, data structures
have \key{constructors}.

\begin{definition}

A constructor is a modification method that resets a data structure to some
particular initial value.

\end{definition}

If we have another instruction that resets all the bits of word of a given word
to $0$, and let \ref{word-0} represent an empty spot, we are able to
consistently model the real world occupation of the parking lot.

% A program is stored in memory, and therefore a particular word in memory can
% always be initialized to some particular value.

Using a 32-bit word for a boolean is not particularly efficient. It can
represent $2^{32}$ different values and we're only utilizing two\footnotemark!
Another use-case for words is to represent natural numbers in the range
$[0;2^{32})$.  Useful basic methods of modification would be addition,
subtraction, multiplication, etc.

\footnotetext{Sometimes it proves practical to waste a little space in favor of
much faster execution time, so while we could do with a single bit to represent
a boolean value, computers often use 1, 2, or even 4 bytes due to the hardware
design.}

As a matter of fact, the common computer comes with these and a myriad of other
instructions, which hence make up the computer's \key{instruction set}. The
``word'' in this context becomes a rather ubiquitous data structure. Most of
the time we won't be needing this ubiquity, and may indeed choose only a small
subset of the instruction set to apply to any particular word in memory.
Indeed, we may desire for a particular word to represent a boolean or a natural
number, but not both at the same time!

While computers could probably support a few more basic data types, it has
proven to be more practical to facilitate this sort of ``safety'' at a
different level of abstraction. Enter \key{types}.

\begin{definition}

A data structure is a type. A value of a given type may only be modified using
a method defined in the corresponding data structure.

\end{definition}

This definition may seem indistinguishible from the definition of a data
structure, but the important distinction is that a data structure has methods
of modification that maintain its ordering, and here we prohibit the use of any
other methods for the data structure.

Imperative programming languages facilitate the definition of types, and like
computers have basic data structures, have basic types. For a given computer
with the basic data structure of a 4-byte word, the basic types may be subsets
or supersets of a word, all depending on the flexibility of the intruction set
of the computer. Hence, some common basic types are presented in
\referToTable{basic-types}.

\makeTable
{basic-types}
{Basic Types}
{lll}
{name & bytes & description}
{
  \code{int} & $4$ & natural number in the range $[-2$
}



 By far the most programs won't be needing the entire
instruction set of a modern computer,

 but the intention of the hardware industry
is to support as many programs as possible, rather than any single one.


Because maintaining a data structure generally requires a sequence of actions..
blah.

A programming language facilitates a programming paradigm if it is both easy,
and reasonably safe to use the paradigm. Having the ability to define data
structures, and actions that modify them, is one thing.  Another is having
assurance that the wrong action is never used to modify the wrong data
structure. This is where \key{type systems} come in handy.

We say that a set of data structures defining the same ordering on the same
\key{types} of data elements, and having the same methods of modification,
belong to the same \key{class} of data structures. A class of data structures
is a type of data element, or simply, type. This notion is intentionally
recursive.

We hence say that actions are defined on particular types, and a type system
can hence ensure that the wrong action is not used to modify the wrong data
structure before the program is ever run. You should by now have plenty of
experience with the notorious SML type system.

% It is clear that designing a new piece of hardware for every type of action is
% impractical. The desired net effect can often be achieved through a series of
% basic actions. For instance, the sum of $n$ different numbers can be computed
% using $n-1$ binary additions. Hence, we can make due with a \key{sequencing
% mechanism} and good old binary addition, rather than design a new piece of
% hardware for every conceivable $n$.

% A computer, therefore, has a finite set of basic actions, called its
% \key{instruction set}. A computer program is specified in terms of a sequence
% of these basic actions, called a \key{program text}. The computer executes the
% program by performing the sequence of actions in order.

% The sequence of basic actions is not a sequencing mechanism in and of itself.
% Indeed, we are unable to specify a computer program for summing a sequence of
% an arbitrary size $n$, since no sequence of binary additions will ever be
% enough for every conceivable $n$.  What's more, why duplicate the actions?
% Afterall, as we sequentially sum a sequence of numbers we're doing the same
% action, a binary addition, just on different data.

% To mitigate this lack of flexibility, the von Neuman architecture introduces
% the \key{jump} action. Since the program is specified as a finite sequence
% of basic actions, we can easily assign a unique integer index to every action in the sequence.
% As the computer is performing an action at a particular index at any point in time

% The
% idea is then to let the computer have an action that allows to specify an arbitrary action as
% the next action to perform, rather then the one then immediately next
% instruction.

% At this point it makes sense to reconsider the concept of time. The convention is to let all basic
% actions be upper bounded by some constant value. This allows us to analyze the
% running time of computer programs in terms of how much sequencing is going on.


% At any point in time, the computer is performing some particular action, 

% and provide a
% mechanism for performing a sequence of actions in order. This however, still does not allow us to write a 

% Depending on our level of interest, we may consider an action as either an
% atomic happening, or as a sequence of subactions necessary to achieve the
% overall net effect. The subactions themselves may be considered in terms of
% their subsubactions, and so on. Indeed, one possible program design method is
% to start with a high-level overview and to gradually dig into the details.
% Naturally, such a hierarchical ordering of a program also facilitates the human
% comprehension of the program. This calls to attention a quote by Donald
% Knuth\cite{knuth-review-of-sp}:

% \begin{quote}"You have perhaps had a dream much like mine: Wouldn't it be nice
% to have a glorious system of complete maps of the world, whereby one could (by
% turning dials) increase or decrease the scale at will? A similar thing can be
% achieved for programs, rather easily, when we give the programs a hierarchic
% structure like those constructed step-wise. It is not hard to imagine a
% computing system which displays a program in such a way that, by pressing
% an-appropriate button, one can replace the name of a routine by its expansion
% on the next level, or conversely."\end{quote}

% Here the word ``routine'' stands for an action, and ``its expansion on the next
% level'' is the sequence of sub-actions necessary to achive the intended
% net effect. The choice of the word ``routine'' is not an unlucky one. Afterall,
% we all have some sort of routine for doing the laundry, and doing the laundry
% is a rather routine task! Let us now consider the immediate expansion of doing
% the laundry, assuming that we have a common washing machine at our disposal:

The dirty pile of laundry is not a particularly interesting data structure in
and of itself, but it will prove interesting to consider how we morph the data
structure into other well-kept data structures as we do the laundry.

Depending on our level of interest, we may consider an action as an atomic
happening or as a sequence of subactions necessary to achieve the overall net
effect. Each subaction can in turn be considered in terms of its subsubactions,
and so on. Let us consider the possible subactions of doing the laundry,
assuming that we have a common washing machine at our disposal:

\begin{codebox}
\li Split the pile into colors and whites.
\li Wash colors.
\li Wash whites.
\li Make a neat stack of the clean laundry.
\end{codebox}

This is an example of an \key{algorithm}. An algorithm is a sequence of actions
that if performed in order achieves some intended net effect. Formally
speaking, an algorithm has some sort of input, which it transforms into some
sort of output through a sequence of actions. Here, the input is a pile of
dirty laundry, and the output is a neat stack of clean laundry. When specifying
an algorithm we tend to give it a name, and be somewhat more explicit about its
inputs and outputs, for instance:

\begin{codebox}
\Procname{$\proc{Do-The-Laundry}\p{\text{A dirty pile of laundry}}$}
\li Split the pile into colors and whites.
\li Wash colors.
\li Wash whites.
\li Make a neat stack of the clean laundry.
\li \Return the neat stack of clean laundry.
\end{codebox}

We all know that whites and colors don't mix well in a washing machine. Indeed,
a higher temperature should be used for whites, since they are generally harder
to get clean, and a lower temperature should be used for colors, since high
temperatures may leed to color bleeding.

A common mistake is to oversee a colored piece in a pile of whites, and wash
the pile with high temperature.  Most likely, this will lead to all whites in
that pile bleeding from white to a light shade of the mixed-in color. That
would definitely cause a hurricane on the other side of the globe!  As such,
this is a \key{pre-} and \key{postcondition} pair we forgot to mention. Not
only should the pile be clean, but every piece of laundry should also keep its
original color.

\begin{definition}

Assuming that preconditions hold before the action commences, the
postconditions should hold after the action terminates\footnotemark.

\footnotetext{We say ``should'', becuase there is currently no automatic way of
assuring this for an arbitrary action in by far the most practical programming
languages. Hence, it is up to the programmer to manually prove that the
postconditions will hold.}

\end{definition}

We now state the more accurate pre- and postconditions of doing the laundry:

\begin{codebox}
\zi
$\begin{array}{ll}
&\text{Given a dirty pile, the pile is cleaned.}\\
\wedge&\text{Every piece of laundry keeps its original color.}
\end{array}$
\end{codebox}

Here, the symbol $\wedge$ means ``and'', i.e. both statements should hold.

When \key{declaring} an action, we also specify its name, as well input and
output parameters. So a complete declaration of doing the laundry looks like
this:

\begin{codebox}
\Procname{$\proc{Do-The-Laundry}::(Laundry)\rightarrow Laundry$}
\zi
$\begin{array}{ll}
&\text{Given a dirty pile, a clean pile is returned.}\\
\wedge&\text{Every piece of laundry keeps its original color.}
\end{array}$
\end{codebox}



as well as the pre- and postconditions. When \key{defining}
an action, we specify its actual sequence of subactions. We make this
distinction because an action declaration may well have several definitions.

Of course, at some point, some real work has to get done.

A von-Neuman computer has a basic set of actions that it can perform.


The initial pile of dirty laundry is not a particularly interesting data
structure, but it will prove beneficial to morph this data structure into other
types as we \emph{progress} through the action. Indeed, 

Of course, at some point we must reach basic, non-divisible 


 i.e.  doing
the laundry, or as a sequence of subactions. This sequence represents the steps
necessary to achieve the overall intended net effect.


(Naur,
function vs. process).

\begin{codebox}
\Procname{$\proc{Do-The-Laundry}\p{dirtyPile}$}
\li \Return $cleanPile$
\end{codebox}

We can dissect this definition into
\key{pre-conditions} and \key{post-conditions}, and state their relationship in
terms of an \key{implication}, denoted by the symbol $\Rightarrow$:

\begin{equation}
\text{Dirty pile}\Rightarrow \text{Clean pile}.
\end{equation}



More formally, if at time $T_n$ we
have a pile of dirty laundry, and perform an action in the subsequent time slot
$T_1$, then at time $T_{n+1}$ we have a pile of clean laundry.


To define the
net effect of an action, we state the conditions that must hold at time
$T_{n+1}$, given that certain conditions hold at time $T_n$. We call these
conditions the \key{post-} and \key{pre-conditions}, respectively. Given an
action $a$, we say that if the preconditions hold, then after the action, the
post-conditions must hold, denoted

\begin{equation}
\text{Precondition} \Rightarrow^a \text{Postcondition}.
\end{equation}


The effect
of the action is that a precondition \emph{implies} a post condition, we denote
this with the symbol $\Rightarrow$.

It does not make sense to do the laundry without a pile of dirty laundry, and a
good ``doing the laundry'' action leaves all the laundry clean after the action
took place. We therefore state the following:

 Before the action
commences, you have a pile of dirty laundry; after the action, the laundry is
clean. We call these the \key{pre-} and \key{postconditions} of an action,
respectively. In particular, it does not make sense to do the laundry unless
you have a pile of dirty laundry, and the post-condition that the laundry is
clean is probably useful for future actions.

The post-conditions are not complete. There are certain things that we would
like to do to facilitate further actions. For instance, it is useful if in the
process of doing the laundry we didn't mix clean and dirty laundry, and at best
-- neatly stacked the clean laundry in the closet.


The post-conditions are not
completely well-formed however. In particular, 

where whites are mixed with
colors. Anyone who's ever done their laundry knows that it's no good to mix 

After the action, the clothes should be neatly stacked in the closet. 

% doing the laundry 
